Estimating the Initial Susceptible Fraction in Dengue Dynamics

### 2014-09-22 draft_lm1.tex
working on the model section

In the particular case of Dengue fever in Brasil, with four circulating 
serotypes, it can be very challenging to parameterize a simulation starting at 
any time other than when the date the disease was reintroduced in the country 
in the late 1980s. For SIR-like models, the relevant initial values are the 
number of infectious and susceptible  individuals. The number of infectious 
individuals are less challenging to determine, because the get very small 
during the winters and can be derived from incidence data  with is monitored 
continuously. The real challenge is to determine the number of susceptibles to 
a particular strain right at the beginning of the dengue season.

Multiple factors contribute to the dificulty in estimating the number of 
susceptibles in the population, first and foremost is the fact that dengue 
confers permanente immunity within serotypes but with only short-lived 
cross-immunity. Secondly, there are a larger number of asymptomatic cases in 
any epidemic, which nevertheless acquire immunity. Lastly Demography and 
migrations affect the number of susceptible in ways which are not easy to 
determine.

The initial number of susceptibles -- $S_0$ -- is a a very powerful parameter 
with the regards to epidemic risk. Its value affects directly the local $R_0$, 
thus being intrumental in predicting epidemics, but it also affects the speed 
and size of the epidemic.

In this paper, we estimate the initial number of susceptibles right before 
every epidemic in the last 18 years. For this we use a simplified model of 
dengue transmission in which the vector is left out. We also model the mix of 
circulating serotypes without discriminating between them. We then fit the 
model output to observed data to estimate the $S_0$ before every 
recorded epidemic in the Brazilian city of Rio de Janeiro. Rio de Janeiro has 
bee through many Dengue epidemics of various serotypes and sizes.

\section*{Methods}
\subsection*{The data}
The Data used to fit the model consists of weekly incidence from 1996 to 2014 
in the city of Rio de Janeiro, Brazil. From this data, we calculated series 
effective reproduction number, $R_t$, time series using the method described by 
Nishiura et al.\cite{nishiura}.\todo[inline]{Add here more information about 
the calculation of Rt such as generation time used, moving windows, etc}
\subsection*{The model}
\subsection*{Parameter estimation using Bayesian melding}
% Results and Discussion can be combined.
\section*{Results}
% We only support three levels of headings, please do not create a heading level below \subsubsection.
%\subsection*{Subsection 1}
\section*{Discussion}
% Do NOT remove this, even if you are not including acknowledgments.
\section*{Acknowledgments}
LMC is grateful to Dr. Leonardo Bastos for useful discussions on the estimation of $R(t)$.

>>>>

In the particular case of Dengue fever in Brazil, with four circulating 
serotypes, it can be very challenging to parameterize a simulation starting at 
any time other than when the date the disease was reintroduced in the country 
in the late 1980s.
For SIR-like models, the relevant initial values are the 
number of infectious and susceptible  individuals. 
The number of infectious 
individuals are less challenging to determine, because the get very small 
during the winters and can be derived from incidence data  with is monitored 
continuously.
The real challenge is to determine the number of susceptibles to 
a particular strain right at the beginning of the dengue season.

Multiple factors contribute to the dificulty in estimating the number of 
susceptibles in the population, first and foremost is the fact that dengue 
confers permanent immunity within serotypes but with only short-lived 
cross-immunity.
Secondly, there are a larger number of asymptomatic cases in 
any epidemic, which nevertheless acquire immunity.
Lastly Demography and 
migrations affect the number of susceptible in ways which are not easy to 
determine.

The initial number of susceptibles -- $S_0$ -- is a a very powerful parameter 
with regard to epidemic risk.
Its value affects directly the local $R_0$, 
thus being intrumental in predicting epidemics, but it also affects the speed 
and size of the epidemic.

In this paper, we estimate the initial number of susceptibles right before 
every epidemic in the last 18 years.
For this we use a simplified model of 
dengue transmission based on a Susceptible-infecctious-Removed (SIR) model.
We then fit the 
model output to observed data to estimate the $S_0$ before every 
recorded epidemic in the Brazilian city of Rio de Janeiro. Rio de Janeiro has 
bee through many Dengue epidemics of various serotypes and sizes.

\section*{Methods}
\subsection*{The data}

The data used to fit the model consists of weekly incidence from 1996 to 2014 
in the city of Rio de Janeiro, Brazil.
From this series, we calculated the effective reproduction number, $R_t$, according to the expression given by Nishiura et al.~\cite{nishiura}.
The method assumes the disease counts $Y(t)$ are Poisson random variables such that $R_t$ is 
\[ R_t = \left(\frac{Y(t + 1)}{Y(t)}\right)^{1/n}\]
where $n$ is the ratio between the length of reporting interval and the mean generation time of the disease.
In this paper we obtain the estimates of $R_t$ using a moving window of three weeks, which roughly corresponds to the generation time of dengue, and thus $n = 1$.
We provide more detail on the estimation and confidence/credibility intervals in Text S1.

\subsection*{The model}

A Susceptible-Infectious-Removed (SIR) model is proposed to model dengue dynamics.
In the traditional formulation of the model, transmission is governed by a transmission rate $\beta$ and recovery happens at a rate $\tau$.
As the epidemic progresses, the effective transmission  rate changes and is
\begin{equation}
 \label{eq:effbeta}
 \beta(t) = R_t	\cdot\tau
\end{equation}

We parameterize our model with the varying transmission using the system of ordinary differential equations
\begin{align}
 \frac{dS}{dt} &= -\beta(t)SI \\
 \frac{dI}{dt} &= \beta(t)SI - \tau I&\\
 \frac{dR}{dt} &= \tau I&
\end{align}
where $S + I + R = 1 \: \forall\: t$. % \in [T_0, T_1]$. 
This is a rather simplified model, in which, for instance, the vector is left out.
Also, although there probably are multiple circulating serotypes, our approach does not discriminate between them.

\subsection*{Bayesian parameter estimation}
For the fitting procedure the time series were normalized to lie on the $[0,1]$ interval, therefore avoiding scaling issues.
% Results and Discussion can be combined.xx
\section*{Results}
% We only support three levels of headings, please do not create a heading level below \subsubsection.
%\subsection*{Subsection 1}
\section*{Discussion}
% Do NOT remove this, even if you are not including acknowledgments.
\section*{Acknowledgments}
LMC is grateful to Dr. Leonardo Bastos for useful discussions on the estimation of $R(t)$.
